Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-13 manuscript/manuscript.tex
adding TICA text (#136)

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
and a transition matrix $P(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
Let us make two remarks to avoid common misconceptions:
\begin{enumerate}
\item Equilibrium:
While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
they are useful to analyze short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
\item Markovianity:
An MSM is a memoryless model.
Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
\end{enumerate}

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i p_{ij} = \pi_j p_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
The CK property for a Markovian matrix is,
\begin{equation}
P(k \tau) = P^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues.
The highest eigenvalue, $\lambda_1(\tau)$, is unique and equal to $1$.
Its corresponding left eigenvector is the stationary distribution, $\bm{\pi}$:
\begin{equation}
\bm{\pi}^\top  P(\tau) = \bm{\pi}^\top.
\end{equation}

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
\begin{equation}
P(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
\end{equation}
where the eigenvalue-eigenvector pairs are indexed in decreasing order.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.

@@

>>>>

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
and a transition matrix $\mathbf{P}(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
Let us make two remarks to avoid common misconceptions:
\begin{enumerate}
\item Equilibrium:
While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
they are useful to analyze short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
\item Markovianity:
An MSM is a memoryless model.
Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
\end{enumerate}

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}$, which is defined for the specified lag time.
For MD simulations in equilibrium, $\mathbf{P}$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i p_{ij} = \pi_j p_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
The CK property for a Markovian matrix is,
\begin{equation}
\mathbf{P}(k \tau) = \mathbf{P}^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues.
The highest eigenvalue, $\lambda_1(\tau)$, is unique and equal to $1$.
Its corresponding left eigenvector is the stationary distribution, $\bm{\pi}$:
\begin{equation}
\bm{\pi}^\top  \mathbf{P}(\tau) = \bm{\pi}^\top.
\end{equation}

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
\begin{equation}
\mathbf{P}(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
\end{equation}
where the eigenvalue-eigenvector pairs are indexed in decreasing order.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.

@@

Our recommended method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
Then, we can solve the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are in descending order.

Dimensionality reduction is achieved by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}=[\mathbf{u}_1 \dots \mathbf{u}_d]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
while retaining the kinetic variance
\begin{equation}
\textrm{KV}_d = \sum\limits_{i=1}^d \lambda_i^2(\tau);
\end{equation}
the total kinetic variance is the sum of the squares of all eigenvalues.
